---
title: "Bennedsen, Hillebrand, Koopman, and Larsen (2024)"
author: "Eric Hillebrand"
date: "04/07/2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

[https://arxiv.org/abs/2404.05401](https://arxiv.org/abs/2404.05401)

## State space methods for $\delta^{18}O$ and $\delta^{13}C$

In the [paper](https://arxiv.org/abs/2404.05401), we specify random walk plus noise models and integrated random walk plus noise models for the time series of $\delta^{18}$O and $\delta^{13}$C measurements from benthic foraminifera collated in [Westerhold et al. (2020)](https://www.science.org/doi/full/10.1126/science.aba6853?). This R-Markdown script contains estimation code for all models discussed in the paper. It also provides code that imputes equidistantly time-stamped smoothed values (in the sense of the Kalman smoothing recursions) for a chosen model and a chosen (fixed) time increment.

We provide download links for regular smoothed values from the preferred model for a number of fixed time increments at the end of this document. This R-Markdown script will become downloadable upon publication of the paper, so that it can serve for replication and for imputing regularly time-stamped imputed values from any model and for any time increment.

The data plot below shows the time series of $\delta^{(18)}$O (left panel), $\delta^{13}$C (middle panel), and the time distance between consecutive observations. It is apparent that the later part of the record is denser than the early part.



```{r, data preparation and time series plots, fig.show="hold", out.width="33%", echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(zoo)
library(KFAS)

westerhold_data_unsort<-read.csv(url("https://raw.githubusercontent.com/chisos9/data_public/master/aba6853_tables_s8_s34.csv"),skip=1)
westerhold_data<-westerhold_data_unsort[order(westerhold_data_unsort$age_tuned),]

timeax_iso<-westerhold_data$age_tuned
timeax_iso<- rev(-timeax_iso) # time runs from -66 MYA to 0
d18O<- westerhold_data$benthic.d18O.VPDB.CorrAdjusted
d18O<-rev(d18O) # Delta 18O runs from -66 MYA to 0, but is not multiplied by (-1)
d13C<-westerhold_data$benthic.d13C.VPDB.CorrAdjusted
d13C<-rev(d13C)

cT_iso<-length(timeax_iso)
idx_axis<-(1:cT_iso) # the original time axis has some multiple observations at 599 points, some are double, few are triple, one is quadruple

# Studies
vStudies<-rev(westerhold_data$benthic.Source) # the studies that provide the data in same order as d18O
vSpecies<-rev(westerhold_data$benthic.Species)
unique_studies<-unique(vStudies)
unique_species<-unique(vSpecies)

# Organization
# [1] "Barnet et al. 2017"               "Hull et al. 2020"                 "Barnet et al. 2019"              
# [4] "Littler et al. 2014"              "McCarren et al. 2008"             
# [6] "Lauretano et al. 2015"            "Stap et al. 2010"                 "Thomas et al. 2018"              
# [9] "Lauretano et al. 2016; 2018"      "Sexton et al. 2011"               "this study"                      
# [12] "Westerhold et al. 2018 LLTM"      "Boscolo Galazzo et al. 2014"      "Lear et al. 2004"                
# [15] "Coxall et al. 2005"               "Coxall & Wilson, 2011"            "Riesselmann et al. 2007"         
# [18] "Paelike2006 PI Herrle"            "Wade and Paelike 2004"            "Liebrand et al. 2011;2016"       
# [21] "Holbourn et al. 2015"             "Tian et al. 2014"                 "Holbourn et al. 2014"            
# [24] "Holbourn et al. 2007;2013;2018"   "Drury et al. 2017"                "Tian et al. 2018"                
# [27] "UniCambridge Hodell this study"   "Bell et al. 2014"                 "MARUM Bremen this study"         
# [30] "Franz & Tiedemann2002"            "Tiedemann & Franz 1997"           "Billups et al. 1998"             
# [33] "Bickert et al. 1997               "deMenocal et al. 1997" 
# [35] Q

vVarPos<-matrix(NA,cT_iso,1)
for (i in 1:cT_iso){ # assign the source study to each point in time
  if (vStudies[i]==unique_studies[1]){
    vVarPos[i]<-1
  } else if (vStudies[i]==unique_studies[2]){
    vVarPos[i]<-2
  } else if (vStudies[i]==unique_studies[3]){
    vVarPos[i]<-3
  } else if (vStudies[i]==unique_studies[4]){
    vVarPos[i]<-4
  } else if (vStudies[i]==unique_studies[5] || vStudies[i]==unique_studies[6]){
    vVarPos[i]<-5
  } else if (vStudies[i]==unique_studies[7]){
    vVarPos[i]<-6
  } else if (vStudies[i]==unique_studies[8]){
    vVarPos[i]<-7
  } else if (vStudies[i]==unique_studies[9]){
    vVarPos[i]<-8
  } else if (vStudies[i]==unique_studies[10]){
    vVarPos[i]<-9
  } else if (vStudies[i]==unique_studies[11]){
    vVarPos[i]<-10
  } else if (vStudies[i]==unique_studies[12]){
    vVarPos[i]<-11
  } else if (vStudies[i]==unique_studies[13]){
    vVarPos[i]<-12
  } else if (vStudies[i]==unique_studies[14]){
    vVarPos[i]<-13
  } else if (vStudies[i]==unique_studies[15]){
    vVarPos[i]<-14
  } else if (vStudies[i]==unique_studies[16]){
    vVarPos[i]<-15
  } else if (vStudies[i]==unique_studies[17]){
    vVarPos[i]<-16
  } else if (vStudies[i]==unique_studies[18]){
    vVarPos[i]<-17
  } else if (vStudies[i]==unique_studies[19]){
    vVarPos[i]<-18
  } else if (vStudies[i]==unique_studies[20]){
    vVarPos[i]<-19
  } else if (vStudies[i]==unique_studies[21]){
    vVarPos[i]<-20
  } else if (vStudies[i]==unique_studies[22]){
    vVarPos[i]<-21
  } else if (vStudies[i]==unique_studies[23]){
    vVarPos[i]<-22
  } else if (vStudies[i]==unique_studies[24]){
    vVarPos[i]<-23
  } else if (vStudies[i]==unique_studies[25]){
    vVarPos[i]<-24
  } else if (vStudies[i]==unique_studies[26]){
    vVarPos[i]<-25
  } else if (vStudies[i]==unique_studies[27]){
    vVarPos[i]<-26
  } else if (vStudies[i]==unique_studies[28]){
    vVarPos[i]<-27
  } else if (vStudies[i]==unique_studies[29]){
    vVarPos[i]<-28
  } else if (vStudies[i]==unique_studies[30]){
    vVarPos[i]<-29
  } else if (vStudies[i]==unique_studies[31]){
    vVarPos[i]<-30
  } else if (vStudies[i]==unique_studies[32]){
    vVarPos[i]<-31
  } else if (vStudies[i]==unique_studies[33]){
    vVarPos[i]<-32
  } else if (vStudies[i]==unique_studies[34] || vStudies[i]==unique_studies[35]){
    vVarPos[i]<-33
  } else if (vStudies[i]==unique_studies[36]){
    vVarPos[i]<-34
  }
}

# > unique_species
# [1] "NTRUE"                                         "OUMB"                                         
# [3] "CSUBS or CEOCEA, not specified in publication" "CPRAE"                                        
# [5] "CSPP"                                          
# [6] "CSPP, whole specimen"  (25 obs)                       
# [7] "CSPP, >250"            (353 obs)                        
# [8] "CSPP, >250, Reruns"    (58 obs)                       
# [9] "CSPP, specimen >250 µm" (10 obs)                        
# [10] "CSPP, 150-250"        (2 obs)  
# [6-10] merged into [7]                        
# [11] "CGRIM"                                         "CHAVA"                                        
# [13] "CMUND"                                         "PWUEL; CMUND"                                 
# [15] "PWUEL (eitherCKULL or PWUEL)"                  "PWUEL"                                        
# [17] "CKULL"                                         "UVIG"                                         
# [19] "GORB" (0 obs)                                 
# [20] "PMURR" (0 obs)                                        
# [21] "NUMB"                                          "CBRA"                                         
# [23] "CCIC" 

vVarPosSpecies<-matrix(NA,cT_iso,1)
for (i in 1:cT_iso){ # assign the source study to each point in time
  if (vSpecies[i]==unique_species[1]){
    vVarPosSpecies[i]<-1
  } else if (vSpecies[i]==unique_species[2]){
    vVarPosSpecies[i]<-2
  } else if (vSpecies[i]==unique_species[3]){
    vVarPosSpecies[i]<-3
  } else if (vSpecies[i]==unique_species[4]){
    vVarPosSpecies[i]<-4
  } else if (vSpecies[i]==unique_species[5]){
    vVarPosSpecies[i]<-5
  } else if (vSpecies[i]==unique_species[6]){
    vVarPosSpecies[i]<-6
  } else if (vSpecies[i]==unique_species[7]){
    vVarPosSpecies[i]<-7
  } else if (vSpecies[i]==unique_species[8]){
    vVarPosSpecies[i]<-8
  } else if (vSpecies[i]==unique_species[9]){
    vVarPosSpecies[i]<-9
  } else if (vSpecies[i]==unique_species[10]){
    vVarPosSpecies[i]<-10
  } else if (vSpecies[i]==unique_species[11]){
    vVarPosSpecies[i]<-11
  } else if (vSpecies[i]==unique_species[12]){
    vVarPosSpecies[i]<-12
  } else if (vSpecies[i]==unique_species[13]){
    vVarPosSpecies[i]<-13
  } else if (vSpecies[i]==unique_species[14]){
    vVarPosSpecies[i]<-14
  } else if (vSpecies[i]==unique_species[15]){
    vVarPosSpecies[i]<-15
  } else if (vSpecies[i]==unique_species[16]){
    vVarPosSpecies[i]<-16
  } else if (vSpecies[i]==unique_species[17]){
    vVarPosSpecies[i]<-17
  } else if (vSpecies[i]==unique_species[18]){
    vVarPosSpecies[i]<-18
  } else if (vSpecies[i]==unique_species[19]){
    vVarPosSpecies[i]<-19
  } else if (vSpecies[i]==unique_species[20]){
    vVarPosSpecies[i]<-20
  } else if (vSpecies[i]==unique_species[21]){
    vVarPosSpecies[i]<-21
  } else if (vSpecies[i]==unique_species[22]){
    vVarPosSpecies[i]<-22
  } else if (vSpecies[i]==unique_species[23]){
    vVarPosSpecies[i]<-23
  }
}

Dtau<-c(-99,diff(timeax_iso)) # version for finding positions: Delta_{tau}=t_{tau}-t_{tau-1} is at tau
Dtau_SSM<-c(diff(timeax_iso),NA) # version for SSM state equation: Delta_{tau+1}=t_{tau+1}-t_tau is at tau

timeax_unique<-unique(timeax_iso)
cT<-length(timeax_unique)
cT_mult<-length(timeax_iso)-cT
timeax_rle<-rle(timeax_iso) # Run length encoding, corresp to unix "uniq -c", returns
                            # list with (no of lengths of each run) and (values)
                            # max(timeax_rle$lengths) gives max number of multiples



# main data matrix mData organization
# columns (the max number of multiple obs at one time stamp is 4)
# 1   | 2               | 3               | 4         | 5         | 6       | 7       | 8       | 9       | 10      | 11      | 12      | 13      | 14        | 15        | 16        | 17        | 18        | 19        | 20        | 21        | 22          | 23          | 24          | 25          | 26          | 27          | 28          | 29                                                                 
# Obs | Obs timeax_iso  | Obs Westerhold  | age_tuned | Delta age | d18O(1) | d18O(2) | d18O(3) | d18O(4) | d13C(1) | d13C(2) | d13C(3) | d13C(4) | 18OStudy1 | 18OStudy2 | 18OStudy3 | 18OStudy4 | 13CStudy1 | 13CStudy2 | 13CStudy3 | 13CStudy4 | 18OSpecies1 | 18OSpecies2 | 18OSpecies3 | 18OSpecies4 | 13CSpecies1 | 13CSpecies2 | 13CSpecies3 | 13CSpecies4
mData<-matrix(NA,cT,29) # records observations and is stored for use in model estimation
mData_multiples<-matrix(NA,cT_mult,29) # only contains multiple obs at one time stamp
k<-1 # will go from 1:cT_iso
l<-1 # will go from 1:cT_mult
for (i in 1:cT){
    if (timeax_rle$lengths[i]==1){
      mData[i,1]<-i # final obs no
      mData[i,2]<-k # index in timeax_iso
      mData[i,3]<-cT_iso-k+1 # original obs no (in Westerhold data)
      mData[i,4]<-timeax_unique[i] # time stamp
      mData[i,5]<-Dtau_SSM[k] # delta t
      mData[i,6]<-d18O[k]
      if (!is.na(d18O[k]))
      {
        mData[i,14]<-vVarPos[k] # study identifier
        mData[i,22]<-vVarPosSpecies[k] # species identifier
      }
      mData[i,10]<-d13C[k]
      if (!is.na(d13C[k])){
        mData[i,18]<-vVarPos[k] # study identifier
        mData[i,26]<-vVarPosSpecies[k] # species identifier
      }
    } else {
      mData[i,1]<-i # final obs no
      mData[i,2]<-k # index in timeax_iso
      mData[i,3]<-cT_iso-k+1 # original obs no (in Westerhold data)
      mData[i,4]<-timeax_unique[i] # time stamp
      mData[i,5]<-Dtau_SSM[k+timeax_rle$lengths[i]-1] # delta t
      for (j in 1:timeax_rle$lengths[i]){
        mData[i,5+j]<-d18O[k-1+j]
        if (!is.na(d18O[k-1+j])){
          mData[i,13+j]<-vVarPos[k-1+j] # study identifier
          mData[i,21+j]<-vVarPosSpecies[k-1+j] # species identifier
        }
        mData[i,9+j]<-d13C[k-1+j]
        if (!is.na(d13C[k-1+j])){
          mData[i,17+j]<-vVarPos[k-1+j] # study identifier
          mData[i,25+j]<-vVarPosSpecies[k-1+j] # species identifier
        }
      }
      mData_multiples[l,]<-mData[i,]
      l<-l+1
  }
    k<-k+timeax_rle$lengths[i]
}

obsax_plot<-zooreg(mData[,1])
timeax_plot<-zooreg(mData[,4])
d18O_plot<-zooreg(mData[,6])
d13C_plot<-zooreg(mData[,10])
Data<-data.frame(timeax=timeax_plot,d18O=d18O_plot,d13C=d13C_plot)
Dtau_plot<-zooreg(mData[,5])

# time series plot
fig_ts1<-ggplot(d18O_plot, aes(x=timeax_plot,y=d18O_plot)) + geom_line(size=.1, color="blue")
fig_ts1 <- fig_ts1 + ggtitle("Delta-O-18") + xlab("MYA") + ylab(" ") + scale_y_reverse()

fig_ts2<-ggplot(d13C_plot, aes(x=timeax_plot,y=d13C_plot)) + geom_line(size=.1, color="blue")
fig_ts2 <- fig_ts2 + ggtitle("Delta-C-13")  + xlab("MYA") + ylab(" ")

fig_ts3<-ggplot(Dtau_plot, aes(x=obsax_plot,y=Dtau_plot)) + geom_line(size=.2, color="black")
fig_ts3 <- fig_ts3 + ggtitle("Delta t")  + xlab("Observation") + ylab(" ")

# fig_ts3<-ggplot(Data, aes(x=timeax)) + geom_line(aes(y=d18O,color="Delta-O-18"),size=.1) + geom_line(aes(y=d13C,color="Delta-C-13"),size=.1)
# fig_ts3 <- fig_ts3 + ggtitle("Delta ")  + xlab("MYA") + ylab(" ") +   scale_colour_manual("", 
#                       breaks = c("Delta-O-18", "Delta-C-13"),
#                       values = c("black", "blue"))   + theme(
#     legend.position = c(.2, .85),
#     legend.justification = c("right", "bottom"),
#     legend.box.just = "center",
#     legend.margin = margin(6, 6, 6, 6)
#     )

plot(fig_ts1)
plot(fig_ts2)
plot(fig_ts3)
```

## Imputing regularly time stamped values

In the following chunk of R code, you can specify the model you would like to impute smoothed values from by simply setting the value of the variable "Model". The output time series will have equidistant time stamps with constant time increment "Delta" in years. You can set this increment by editing in the code chunk as well. For Model 5, you also need to specify the order $m$ of integration, or, equivalently, the order of the corresponding Butterworth filter in the same fashion. The value set for $m$ is ignored for Models 1--4.

The output time series are saved in a file "output.txt" in the same folder where you have saved this R-Markdown script. The file consists of three columns: (1) time stamps (in (-1) MY format, i.e., starting from -67 and ending at 0), (2) Delta-O-18 values, (3) Delta-C-13 values 

```{r, Settings, echo=TRUE}
Model<-4
Delta<-50000 # years
m<-2 # Integer. Used for Model 5, otherwise ignored. Implemented for m between 2 and 6. Note that m=1 is Model 3.
```



## Model 1: Random walk plus noise

Model 1 specifies the time series as consisting of an unobserved random-walk component plus white noise in the measurement equation:

$$
y_{j,t_\nu} = \mu_{t_\nu} + \varepsilon_{j,t_\nu},
$$
where $t_\nu$, $\nu=1,\ldots, N=23722$, are the time stamps, $j=1,\ldots,4$ are the different measurements of $\delta^{18}O$ or $\delta^{13}C$ at time $t_\nu$ (there are at most 4 observations at each time stamp, and if there are fewer, the corresponding $y_{j,t_\nu}$ are NA), and $\varepsilon_{j,t_\nu}$ is a normal random variable.

The dynamics of the unobserved random walk are specified in the transition equation as

$$
\mu_{t_\nu+\Delta t_\nu} = \mu_{t_\nu} + (\Delta t_\nu)^{\frac{1}{2}}\eta_{t_\nu},
$$
where $\Delta t_\nu = t_{\nu+1}-t_\nu$ and $\eta_{t_\nu}$ is a normal random variable.


```{r, Model 1, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
if (Model==1){
  
  # Delta-O-18
  mObs<-mData[,6:9] 
  mOnes<-matrix(1,4,1)
  mR18<-array(0,c(1,1,cT))
  vDeltaTau<-matrix(0,cT,1)
  DeltaTau<-0
  for (i in 1:cT){
    if (is.na(mData[i,6])){
      DeltaTau<-DeltaTau+mData[i,5]
    } else {
      DeltaTau<-DeltaTau+mData[i,5]
      vDeltaTau[i]<-DeltaTau
      DeltaTau<-0
    }
  }
  mR18[1,1,]<-sqrt(vDeltaTau)
  mR18[1,1,cT]<-0
  mdl<-SSModel(as.matrix(mObs) ~ -1 + SSMcustom(Z=mOnes,T=1,R=mR18,Q=NA,a1=0,P1=0,P1inf=1),H=matrix(NA,4,4))
  
  # objective function of the state space model
  objf <- function(pars,model,estimate=TRUE){
    model$Q[, ,1]<-exp(pars[1])
    model$H[, ,1]<-diag(exp(pars[2]),4)
    if (estimate){
      -logLik(model)
    } else {
      model
    }
  }
  
  # estimation
  github_url<-"https://github.com/chisos9/data_public/raw/master/Delta18O_locallevel_par.rda"
  load(url(github_url)) # load parameter start vector 
  mdl.opt<-optim(vPi,fn=objf,model=mdl,method="BFGS",control=list(trace=4,maxit=10000),hessian=FALSE) 
  vPi<-mdl.opt$par

  # evaluate Kalman filter and smoother in the optimum
  mdl.fit<-objf(vPi, mdl, estimate = FALSE)
  mdl.kfs<-KFS(mdl.fit,filtering=c('state','signal','disturbance'),smoothing=c('state','signal','disturbance'),simplify=FALSE)
  
  # Generate equidistantly spaced data in time
  timeax<-seq(from=-67,to=0,by=(Delta/1e6))
  cT_gen<-length(timeax)
  mData_equi<-matrix(NA,cT_gen,29)
  mData_equi[,4]<-timeax
  mData_merged<-rbind(mData,mData_equi)
  indices_in_timestamps_order<-order(mData_merged[,4])
  mData_gen<-mData_merged[indices_in_timestamps_order,]
  indices_equi<-matrix(NA,cT_gen,1)
  for (i in 1:cT_gen){
    indices_equi[i]<-which(indices_in_timestamps_order==(cT+i))
  }
  Dtau_SSM_gen<-c(diff(mData_gen[,4]),NA)
  vDeltaTau_gen<-matrix(0,cT+cT_gen,1)
  DeltaTau<-0
  for (i in 1:(cT+cT_gen)){
    if (is.na(mData_gen[i,6])){
      DeltaTau<-DeltaTau+Dtau_SSM_gen[i]
    } else {
      DeltaTau<-DeltaTau+Dtau_SSM_gen[i]
      vDeltaTau_gen[i]<-DeltaTau
      DeltaTau<-0
    }
  }
  mR_gen18<-array(0,c(1,1,cT+cT_gen))
  mR_gen18[1,1,]<-sqrt(vDeltaTau_gen)
  mR_gen18[1,1,cT+cT_gen]<-0
  mdl_gen<-SSModel(mData_gen[,6:9] ~ -1 + SSMcustom(Z=mOnes,T=1,R=mR_gen18,Q=NA,a1=0,P1=0,P1inf=1),H=matrix(NA,4,4))
  mdl.fit_gen<-objf(vPi, mdl_gen, estimate = FALSE)
  mdl.kfs_gen<-KFS(mdl.fit_gen,filtering=c('state','signal','disturbance'),smoothing=c('state','signal','disturbance'),simplify=FALSE)
  output_series_d18O<-mdl.kfs_gen$alphahat[indices_equi]
  
  #------------------------------------------------------------------------------------

  # Delta-C-13
  mObs<-mData[,10:13] 
  mR13<-array(0,c(1,1,cT))
  vDeltaTau<-matrix(0,cT,1)
  DeltaTau<-0
  for (i in 1:cT){
    if (is.na(mData[i,10])){
      DeltaTau<-DeltaTau+mData[i,5]
    } else {
      DeltaTau<-DeltaTau+mData[i,5]
      vDeltaTau[i]<-DeltaTau
      DeltaTau<-0
    }
  }
  mR13[1,1,]<-sqrt(vDeltaTau)
  mR13[1,1,cT]<-0
  mdl<-SSModel(as.matrix(mObs) ~ -1 + SSMcustom(Z=mOnes,T=1,R=mR13,Q=NA,a1=0,P1=0,P1inf=1),H=matrix(NA,4,4))
  
  # estimation
  github_url<-"https://github.com/chisos9/data_public/raw/master/Delta13C_locallevel_par.rda"
  load(url(github_url)) # load parameter start vector 
  mdl.opt<-optim(vPi,fn=objf,model=mdl,method="BFGS",control=list(trace=4,maxit=10000),hessian=FALSE) 
  vPi<-mdl.opt$par

  # evaluate Kalman filter and smoother in the optimum
  mdl.fit<-objf(vPi, mdl, estimate = FALSE)
  mdl.kfs<-KFS(mdl.fit,filtering=c('state','signal','disturbance'),smoothing=c('state','signal','disturbance'),simplify=FALSE)
  
  # Generate equidistantly spaced data in time
  vDeltaTau_gen<-matrix(0,cT+cT_gen,1)
  DeltaTau<-0
  for (i in 1:(cT+cT_gen)){
    if (is.na(mData_gen[i,10])){
      DeltaTau<-DeltaTau+Dtau_SSM_gen[i]
    } else {
      DeltaTau<-DeltaTau+Dtau_SSM_gen[i]
      vDeltaTau_gen[i]<-DeltaTau
      DeltaTau<-0
    }
  }
  mR_gen13<-array(0,c(1,1,cT+cT_gen))
  mR_gen13[1,1,]<-sqrt(vDeltaTau_gen)
  mR_gen13[1,1,cT+cT_gen]<-0
  mdl_gen<-SSModel(mData_gen[,10:13] ~ -1 + SSMcustom(Z=mOnes,T=1,R=mR_gen13,Q=NA,a1=0,P1=0,P1inf=1),H=matrix(NA,4,4))
  mdl.fit_gen<-objf(vPi, mdl_gen, estimate = FALSE)
  mdl.kfs_gen<-KFS(mdl.fit_gen,filtering=c('state','signal','disturbance'),smoothing=c('state','signal','disturbance'),simplify=FALSE)
  output_series_d13C<-mdl.kfs_gen$alphahat[indices_equi]
  output_matrix<-cbind(round(timeax,digits=6),output_series_d18O,output_series_d13C)
  write.table(output_matrix,file="output.txt",sep="\t",row.names=FALSE)
  }
```


## Model 2: Random walk plus noise with study-specific variances

Model 2 differentiates the variances of the error terms in the measurement equation according to the study of origin, of which there are 34:

$$
y_{i_j,t_\nu} = \mu_{t_\nu} + \varepsilon_{i_j,t_\nu},
$$
where $i_j=1,\ldots,34$, $j=1,\ldots,4$, is an index that denotes the origin of the $j$-th observation at time $t_\nu$ from study $i$.  The error term $\varepsilon_{i_j,t_\nu}$ is a normal random variable with variance $\sigma_i^2$, $i=1,\ldots, 34$. The dynamics of the unobserved random walk are as in Model 1.


```{r, Model 2, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
if (Model==2){
  
  # Delta-O-18
  
  mObs<-mData[,6:9] # Delta 18O
  mOnes<-matrix(1,4,1)
  mR18<-array(0,c(1,1,cT))
  vDeltaTau<-matrix(0,cT,1)
  DeltaTau<-0
  for (i in 1:cT){
    if (is.na(mData[i,6])){
      DeltaTau<-DeltaTau+mData[i,5]
    } else {
      DeltaTau<-DeltaTau+mData[i,5]
      vDeltaTau[i]<-DeltaTau
      DeltaTau<-0
    }
  }
  mR18[1,1,]<-sqrt(vDeltaTau)
  mR18[1,1,cT]<-0
  mdl<-SSModel(as.matrix(mObs) ~ -1 + SSMcustom(Z=mOnes,T=1,R=mR18,Q=NA,a1=0,P1=0,P1inf=1),H=array(NA,c(4,4,cT)))

  mH<-array(0,c(4,4,cT))
  # objective function of the state space model
  objf <- function(pars,model,estimate=TRUE){
    model$Q[, ,1]<-exp(pars[35])
    for (i in 1:cT){
      idx<-which(!is.na(mData[i,14:17]))
      mmH<-matrix(0,4,4)
      vDiag<-matrix(0,4,1)
      vDiag[idx]<-exp(pars[mData[i,13+idx]])
      mmH[row(mmH)==col(mmH)]<-vDiag
      mH[, ,i]<-mmH
    }
    model$H<-mH
    if (estimate){
      -logLik(model)
    } else {
      model
    }
  }
  
   # estimation
  github_url<-"https://github.com/chisos9/data_public/raw/master/Delta18O_locallevel_studyvariances_par.rda"
  load(url(github_url)) # load parameter start vector 
  mdl.opt<-optim(vPi,fn=objf,model=mdl,method="BFGS",control=list(trace=4,maxit=10000),hessian=FALSE)
  vPi<-mdl.opt$par
  
  # evaluate Kalman filter and smoother in the optimum
  mdl.fit<-objf(vPi, mdl, estimate = FALSE)
  mdl.kfs<-KFS(mdl.fit,filtering=c('state','signal','disturbance'),smoothing=c('state','signal','disturbance'),simplify=FALSE)

  # Generate equidistantly spaced data in time
  timeax<-seq(from=-67,to=0,by=(Delta/1e6))
  cT_gen<-length(timeax)
  mData_equi<-matrix(NA,cT_gen,29)
  mData_equi[,4]<-timeax
  mData_merged<-rbind(mData,mData_equi)
  indices_in_timestamps_order<-order(mData_merged[,4])
  mData_gen<-mData_merged[indices_in_timestamps_order,]
  indices_equi<-matrix(NA,cT_gen,1)
  for (i in 1:cT_gen){
    indices_equi[i]<-which(indices_in_timestamps_order==(cT+i))
  }
  Dtau_SSM_gen<-c(diff(mData_gen[,4]),NA)
  vDeltaTau_gen<-matrix(0,cT+cT_gen,1)
  DeltaTau<-0
  for (i in 1:(cT+cT_gen)){
    if (is.na(mData_gen[i,6])){
      DeltaTau<-DeltaTau+Dtau_SSM_gen[i]
    } else {
      DeltaTau<-DeltaTau+Dtau_SSM_gen[i]
      vDeltaTau_gen[i]<-DeltaTau
      DeltaTau<-0
    }
  }
  mR_gen18<-array(0,c(1,1,cT+cT_gen))
  mR_gen18[1,1,]<-sqrt(vDeltaTau_gen)
  mR_gen18[1,1,cT+cT_gen]<-0
  mH<-array(0,c(4,4,cT+cT_gen))
  
  objf_gen <- function(pars,model,estimate=TRUE){
    model$Q[, ,1]<-exp(pars[35])
    for (i in 1:(cT+cT_gen)){
      idx<-which(!is.na(mData_gen[i,14:17]))
      mmH<-matrix(0,4,4)
      vDiag<-matrix(0,4,1)
      vDiag[idx]<-exp(pars[mData_gen[i,13+idx]])
      mmH[row(mmH)==col(mmH)]<-vDiag
      mH[, ,i]<-mmH
    }
    model$H<-mH
    if (estimate){
      -logLik(model)
    } else {
      model
    }
  }

  mdl_gen<-SSModel(mData_gen[,6:9] ~ -1 + SSMcustom(Z=mOnes,T=1,R=mR_gen18,Q=NA,a1=0,P1=0,P1inf=1),H=array(NA,c(4,4,cT+cT_gen)))
  mdl.fit_gen<-objf_gen(vPi, mdl_gen, estimate = FALSE)
  mdl.kfs_gen<-KFS(mdl.fit_gen,filtering=c('state','signal','disturbance'),smoothing=c('state','signal','disturbance'),simplify=FALSE)
  output_series_d18O<-mdl.kfs_gen$alphahat[indices_equi]
  
  #------------------------------------------------------------------------------------

  # Delta-C-13
  
  mObs<-mData[,10:13] 
  mR13<-array(0,c(1,1,cT))
  vDeltaTau<-matrix(0,cT,1)
  DeltaTau<-0
  for (i in 1:cT){
    if (is.na(mData[i,10])){
      DeltaTau<-DeltaTau+mData[i,5]
    } else {
      DeltaTau<-DeltaTau+mData[i,5]
      vDeltaTau[i]<-DeltaTau
      DeltaTau<-0
    }
  }
  mR13[1,1,]<-sqrt(vDeltaTau)
  mR13[1,1,cT]<-0
  mdl<-SSModel(as.matrix(mObs) ~ -1 + SSMcustom(Z=mOnes,T=1,R=mR13,Q=NA,a1=0,P1=0,P1inf=1),H=array(NA,c(4,4,cT)))
  
  mH<-array(0,c(4,4,cT))
  # objective function of the state space model
  objf <- function(pars,model,estimate=TRUE){
    model$Q[, ,1]<-exp(pars[35])
    for (i in 1:cT){
      idx<-which(!is.na(mData[i,18:21]))
      mmH<-matrix(0,4,4)
      vDiag<-matrix(0,4,1)
      vDiag[idx]<-exp(pars[mData[i,17+idx]])
      mmH[row(mmH)==col(mmH)]<-vDiag
      mH[, ,i]<-mmH
    }
    model$H<-mH
    if (estimate){
      -logLik(model)
    } else {
      model
    }
  }
  
  # estimation
  github_url<-"https://github.com/chisos9/data_public/raw/master/Delta13C_locallevel_studyvariances_par.rda"
  load(url(github_url)) # load parameter start vector 
  mdl.opt<-optim(vPi,fn=objf,model=mdl,method="BFGS",control=list(trace=4,maxit=10000),hessian=FALSE)
  vPi<-mdl.opt$par
  
  # evaluate Kalman filter and smoother in the optimum
  mdl.fit<-objf(vPi, mdl, estimate = FALSE)
  mdl.kfs<-KFS(mdl.fit,filtering=c('state','signal','disturbance'),smoothing=c('state','signal','disturbance'),simplify=FALSE)

  # Generate equidistantly spaced data in time
  vDeltaTau_gen<-matrix(0,cT+cT_gen,1)
  DeltaTau<-0
  for (i in 1:(cT+cT_gen)){
    if (is.na(mData_gen[i,10])){
      DeltaTau<-DeltaTau+Dtau_SSM_gen[i]
    } else {
      DeltaTau<-DeltaTau+Dtau_SSM_gen[i]
      vDeltaTau_gen[i]<-DeltaTau
      DeltaTau<-0
    }
  }
  mR_gen13<-array(0,c(1,1,cT+cT_gen))
  mR_gen13[1,1,]<-sqrt(vDeltaTau_gen)
  mR_gen13[1,1,cT+cT_gen]<-0
  mH<-array(0,c(4,4,cT+cT_gen))
  
  objf_gen <- function(pars,model,estimate=TRUE){
    model$Q[, ,1]<-exp(pars[35])
    for (i in 1:(cT+cT_gen)){
      idx<-which(!is.na(mData_gen[i,18:21]))
      mmH<-matrix(0,4,4)
      vDiag<-matrix(0,4,1)
      vDiag[idx]<-exp(pars[mData_gen[i,17+idx]])
      mmH[row(mmH)==col(mmH)]<-vDiag
      mH[, ,i]<-mmH
    }
    model$H<-mH
    if (estimate){
      -logLik(model)
    } else {
      model
    }
  }
  
  mdl_gen<-SSModel(mData_gen[,10:13] ~ -1 + SSMcustom(Z=mOnes,T=1,R=mR_gen13,Q=NA,a1=0,P1=0,P1inf=1),H=array(NA,c(4,4,cT+cT_gen)))
  mdl.fit_gen<-objf_gen(vPi, mdl_gen, estimate = FALSE)
  mdl.kfs_gen<-KFS(mdl.fit_gen,filtering=c('state','signal','disturbance'),smoothing=c('state','signal','disturbance'),simplify=FALSE)
  output_series_d13C<-mdl.kfs_gen$alphahat[indices_equi]
  output_matrix<-cbind(round(timeax,digits=6),round(output_series_d18O,digits=10),round(output_series_d13C,digits=10))
  write.table(output_matrix,file="output.txt",sep="\t",row.names=FALSE)
  
}
```

## Model 3: Adding climate-state specific variances in the transition equation

Model 3 adds a differentiated variance structure in the transition equation to Model 2. The motivation is the identification of six distinct climate states in the time series record that is reported in [Westerhold et al. (2020)](https://www.science.org/doi/full/10.1126/science.aba6853?). These states are "Warmhouse 2" (67-56 MYA), "Hothouse" (56-47 MYA), "Warmhouse 1" (47-34 MYA), "Coolhouse 1" (34-13.9 MYA), "Coolhouse 2" (13.9-3.3 MYA), "Icehouse" (3.3.0 MYA). The differentiation of these states is captured by changing variances of the white noise driver of the unobserved random walk $\mu$. The measurement equation, with its differentiation of the studies of origin, remains the same as in Model 2. The transition equation becomes
$$
\mu_{t_\nu+\Delta t_\nu} = \mu_{t_\nu} + (\Delta t_\nu)^\frac{1}{2}\eta_{j,t_\nu},\; \eta_{j,t_\nu}\sim\mathsf{N}(0,\sigma_{\eta,j}^2 ),
$$
where $j=1,\ldots,6$ indicates the six different climate states according to whether $t_\nu$ falls into the time intervals mentioned above. 

```{r, Model 3, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
if (Model==3){
  
  # Delta-O-18
  
  mObs<-mData[,6:9]
  mOnes<-matrix(1,4,1)
  mR<-array(0,c(1,1,cT))
  vDeltaTau<-matrix(0,cT,1)
  DeltaTau<-0
  for (i in 1:cT){
    if (is.na(mData[i,6])){
      DeltaTau<-DeltaTau+mData[i,5]
    } else {
      DeltaTau<-DeltaTau+mData[i,5]
      vDeltaTau[i]<-DeltaTau
      DeltaTau<-0
    }
  }
  mR[1,1,]<-sqrt(vDeltaTau)
  mR[1,1,cT]<-0
  mdl<-SSModel(as.matrix(mObs) ~ -1 + SSMcustom(Z=mOnes,T=1,R=mR,Q=array(NA,c(1,1,cT)),a1=0,P1=0,P1inf=1),H=array(NA,c(4,4,cT)))
  
  mH<-array(0,c(4,4,cT))
  mQ<-array(0,c(1,1,cT))
  
  # objective function of the state space model
  objf <- function(pars,model,estimate=TRUE){
    for (i in 1:cT){
      if (mData[i,4]< (-56)){                                           # 2690
        mQ[1,1,i]<-exp(pars[35])
      } else if ((mData[i,4] >= (-56)) && (mData[i,4] < (-47))) {       # 5693
        mQ[1,1,i]<-exp(pars[36])
      } else if ((mData[i,4] >= (-47)) && (mData[i,4] < (-34))) {       # 7471
        mQ[1,1,i]<-exp(pars[37])
      } else if ((mData[i,4] >= (-34)) && (mData[i,4] < (-13.9))) {     # 13933
        mQ[1,1,i]<-exp(pars[38])
      } else if ((mData[i,4] >= (-13.9)) && (mData[i,4] < (-3.3))) {    # 20179
        mQ[1,1,i]<-exp(pars[39])
      } else if (mData[i,4] >= (-3.3)) {
        mQ[1,1,i]<-exp(pars[40])
      }
    }
    model$Q<-mQ
    for (i in 1:cT){
      idx<-which(!is.na(mData[i,14:17]))
      mmH<-matrix(0,4,4)
      vDiag<-matrix(0,4,1)
      vDiag[idx]<-exp(pars[mData[i,13+idx]])
      mmH[row(mmH)==col(mmH)]<-vDiag
      mH[, ,i]<-mmH
    }
    model$H<-mH
    if (estimate){
      -logLik(model)
    } else {
      model
    }
  }
  
  # estimation
  github_url<-"https://github.com/chisos9/data_public/raw/master/Delta18O_locallevel_studies_periods_par.rda"
  load(url(github_url)) # load parameter start vector 
  mdl.opt<-optim(vPi,fn=objf,model=mdl,method="BFGS",control=list(trace=4,maxit=10000),hessian=FALSE) 
  vPi<-mdl.opt$par

  # evaluate Kalman filter and smoother in the optimum
  mdl.fit<-objf(vPi, mdl, estimate = FALSE)
  mdl.kfs<-KFS(mdl.fit,filtering=c('state','signal','disturbance'),smoothing=c('state','signal','disturbance'),simplify=FALSE)
  
  # Generate equidistantly spaced data in time
  timeax<-seq(from=-67,to=0,by=(Delta/1e6))
  cT_gen<-length(timeax)
  mData_equi<-matrix(NA,cT_gen,29)
  mData_equi[,4]<-timeax
  mData_merged<-rbind(mData,mData_equi)
  indices_in_timestamps_order<-order(mData_merged[,4])
  mData_gen<-mData_merged[indices_in_timestamps_order,]
  indices_equi<-matrix(NA,cT_gen,1)
  for (i in 1:cT_gen){
    indices_equi[i]<-which(indices_in_timestamps_order==(cT+i))
  }
  Dtau_SSM_gen<-c(diff(mData_gen[,4]),NA)
  vDeltaTau_gen<-matrix(0,cT+cT_gen,1)
  DeltaTau<-0
  for (i in 1:(cT+cT_gen)){
    if (is.na(mData_gen[i,6])){
      DeltaTau<-DeltaTau+Dtau_SSM_gen[i]
    } else {
      DeltaTau<-DeltaTau+Dtau_SSM_gen[i]
      vDeltaTau_gen[i]<-DeltaTau
      DeltaTau<-0
    }
  }
  mR_gen<-array(0,c(1,1,cT+cT_gen))
  mR_gen[1,1,]<-sqrt(vDeltaTau_gen)
  mR_gen[1,1,cT+cT_gen]<-0
  mH<-array(0,c(4,4,cT+cT_gen))
  mQ<-array(0,c(1,1,cT+cT_gen))
  
  objf_gen <- function(pars,model,estimate=TRUE){
    for (i in 1:(cT+cT_gen)){
      if (mData_gen[i,4]< (-56)){                                             
        mQ[1,1,i]<-exp(pars[35])
      } else if ((mData_gen[i,4] >= (-56)) && (mData_gen[i,4] < (-47))) {       
        mQ[1,1,i]<-exp(pars[36])
      } else if ((mData_gen[i,4] >= (-47)) && (mData_gen[i,4] < (-34))) {       
        mQ[1,1,i]<-exp(pars[37])
      } else if ((mData_gen[i,4] >= (-34)) && (mData_gen[i,4] < (-13.9))) {     
        mQ[1,1,i]<-exp(pars[38])
      } else if ((mData_gen[i,4] >= (-13.9)) && (mData_gen[i,4] < (-3.3))) {  
        mQ[1,1,i]<-exp(pars[39])
      } else if (mData_gen[i,4] >= (-3.3)) {
        mQ[1,1,i]<-exp(pars[40])
      }
    }
    model$Q<-mQ
    for (i in 1:(cT+cT_gen)){
      idx<-which(!is.na(mData_gen[i,14:17]))
      mmH<-matrix(0,4,4)
      vDiag<-matrix(0,4,1)
      vDiag[idx]<-exp(pars[mData_gen[i,13+idx]])
      mmH[row(mmH)==col(mmH)]<-vDiag
      mH[, ,i]<-mmH
    }
    model$H<-mH
    if (estimate){
      -logLik(model)
    } else {
      model
    }
  }
  
  mdl_gen<-SSModel(mData_gen[,6:9] ~ -1 + SSMcustom(Z=mOnes,T=1,R=mR_gen,Q=array(NA,c(1,1,cT+cT_gen)),a1=0,P1=0,P1inf=1),H=array(NA,c(4,4,cT+cT_gen)))
  mdl.fit_gen<-objf_gen(vPi, mdl_gen, estimate = FALSE)
  mdl.kfs_gen<-KFS(mdl.fit_gen,filtering=c('state','signal','disturbance'),smoothing=c('state','signal','disturbance'),simplify=FALSE)
  output_series_d18O<-mdl.kfs_gen$alphahat[indices_equi]
  
  #------------------------------------------------------------------------------------

  # Delta-C-13
  
  mObs<-mData[,10:13] 
  mR<-array(0,c(1,1,cT))
  vDeltaTau<-matrix(0,cT,1)
  DeltaTau<-0
  for (i in 1:cT){
    if (is.na(mData[i,10])){
      DeltaTau<-DeltaTau+mData[i,5]
    } else {
      DeltaTau<-DeltaTau+mData[i,5]
      vDeltaTau[i]<-DeltaTau
      DeltaTau<-0
    }
  }
  mR[1,1,]<-sqrt(vDeltaTau)
  mR[1,1,cT]<-0
  mdl<-SSModel(as.matrix(mObs) ~ -1 + SSMcustom(Z=mOnes,T=1,R=mR,Q=array(NA,c(1,1,cT)),a1=0,P1=0,P1inf=1),H=array(NA,c(4,4,cT)))
  
  mH<-array(0,c(4,4,cT))
  mQ<-array(0,c(1,1,cT))
  
  # objective function of the state space model
  objf <- function(pars,model,estimate=TRUE){
    model$Q[, ,1]<-exp(pars[35])
    for (i in 1:cT){
      if (mData[i,4]< (-56)){
        mQ[1,1,i]<-exp(pars[35])
      } else if ((mData[i,4] >= (-56)) && (mData[i,4] < (-47))) {
        mQ[1,1,i]<-exp(pars[36])
      } else if ((mData[i,4] >= (-47)) && (mData[i,4] < (-34))) {
        mQ[1,1,i]<-exp(pars[37])
      } else if ((mData[i,4] >= (-34)) && (mData[i,4] < (-13.9))) {
        mQ[1,1,i]<-exp(pars[38])
      } else if ((mData[i,4] >= (-13.9)) && (mData[i,4] < (-3.3))) {
        mQ[1,1,i]<-exp(pars[39])
      } else if (mData[i,4] >= (-3.3)) {
        mQ[1,1,i]<-exp(pars[40])
      }
    }
    model$Q<-mQ
    for (i in 1:cT){
      idx<-which(!is.na(mData[i,18:21]))
      mmH<-matrix(0,4,4)
      vDiag<-matrix(0,4,1)
      vDiag[idx]<-exp(pars[mData[i,17+idx]])
      mmH[row(mmH)==col(mmH)]<-vDiag
      mH[, ,i]<-mmH
    }
    model$H<-mH
    if (estimate){
      -logLik(model)
    } else {
      model
    }
  }
  
  # estimation
  github_url<-"https://github.com/chisos9/data_public/raw/master/Delta13C_locallevel_studies_periods_par.rda"
  load(url(github_url)) # load parameter start vector
  mdl.opt<-optim(vPi,fn=objf,model=mdl,method="BFGS",control=list(trace=4,maxit=10000),hessian=FALSE) 
  vPi<-mdl.opt$par
  
  # evaluate Kalman filter and smoother in the optimum
  mdl.fit<-objf(vPi, mdl, estimate = FALSE)
  mdl.kfs<-KFS(mdl.fit,filtering=c('state','signal','disturbance'),smoothing=c('state','signal','disturbance'),simplify=FALSE)
  
  # Generate equidistantly spaced data in time
  vDeltaTau_gen<-matrix(0,cT+cT_gen,1)
  DeltaTau<-0
  for (i in 1:(cT+cT_gen)){
    if (is.na(mData_gen[i,10])){
      DeltaTau<-DeltaTau+Dtau_SSM_gen[i]
    } else {
      DeltaTau<-DeltaTau+Dtau_SSM_gen[i]
      vDeltaTau_gen[i]<-DeltaTau
      DeltaTau<-0
    }
  }
  mR_gen<-array(0,c(1,1,cT+cT_gen))
  mR_gen[1,1,]<-sqrt(vDeltaTau_gen)
  mR_gen[1,1,cT+cT_gen]<-0
  mH<-array(0,c(4,4,cT+cT_gen))
  mQ<-array(0,c(1,1,cT+cT_gen))
  
  objf_gen <- function(pars,model,estimate=TRUE){
    for (i in 1:(cT+cT_gen)){
      if (mData_gen[i,4]< (-56)){                                               
        mQ[1,1,i]<-exp(pars[35])
      } else if ((mData_gen[i,4] >= (-56)) && (mData_gen[i,4] < (-47))) {       
        mQ[1,1,i]<-exp(pars[36])
      } else if ((mData_gen[i,4] >= (-47)) && (mData_gen[i,4] < (-34))) {       
        mQ[1,1,i]<-exp(pars[37])
      } else if ((mData_gen[i,4] >= (-34)) && (mData_gen[i,4] < (-13.9))) {     
        mQ[1,1,i]<-exp(pars[38])
      } else if ((mData_gen[i,4] >= (-13.9)) && (mData_gen[i,4] < (-3.3))) {    
        mQ[1,1,i]<-exp(pars[39])
      } else if (mData_gen[i,4] >= (-3.3)) {
        mQ[1,1,i]<-exp(pars[40])
      }
    }
    model$Q<-mQ
    for (i in 1:(cT+cT_gen)){
      idx<-which(!is.na(mData_gen[i,18:21]))
      mmH<-matrix(0,4,4)
      vDiag<-matrix(0,4,1)
      vDiag[idx]<-exp(pars[mData_gen[i,17+idx]])
      mmH[row(mmH)==col(mmH)]<-vDiag
      mH[, ,i]<-mmH
    }
    model$H<-mH
    if (estimate){
      -logLik(model)
    } else {
      model
    }
  }
  
  mdl_gen<-SSModel(mData_gen[,10:13] ~ -1 + SSMcustom(Z=mOnes,T=1,R=mR_gen,Q=array(NA,c(1,1,cT+cT_gen)),a1=0,P1=0,P1inf=1),H=array(NA,c(4,4,cT+cT_gen)))
  mdl.fit_gen<-objf_gen(vPi, mdl_gen, estimate = FALSE)
  mdl.kfs_gen<-KFS(mdl.fit_gen,filtering=c('state','signal','disturbance'),smoothing=c('state','signal','disturbance'),simplify=FALSE)
  output_series_d13C<-mdl.kfs_gen$alphahat[indices_equi]
  output_matrix<-cbind(round(timeax,digits=6),round(output_series_d18O,digits=10),round(output_series_d13C,digits=10))
  write.table(output_matrix,file="output.txt",sep="\t",row.names=FALSE)

}
```

## Model 4: Bivariate model for $\delta^{18}$O and $\delta^{13}$C with study-specific measurement and climate-state specific transition variances

The measurement equation of Model 4 is
$$
\begin{align*}
y_{i_j,t_\nu}^{\delta^{18}O} &= \mu_{t_\nu}^{\delta^{18}O} + \varepsilon_{i_j,t_\nu}^{\delta^{18}O},\\
y_{i_j,t_\nu}^{\delta^{13}C} &= \mu_{t_\nu}^{\delta^{13}C} + \varepsilon_{i_j,t_\nu}^{\delta^{13}C},\\
\end{align*}
$$
where $i_j=1,\ldots,34$, $j=1,\ldots,4$, is an index that denotes the origin of the $j$-th observation at time $t_\nu$ from study $i$.  The error term $\varepsilon_{i_j,t_\nu}^{(\delta^{18}O,\delta^{13}C)}$ is a normal random variable with variance $\sigma_{i,(\delta^{18}O,\delta^{13}C)}^2$, $i=1,\ldots, 34$. 

The transition equation for the unobserved component $\mu$ is given as follows.
$$
\begin{align*}
\mu_{t_\nu+\Delta t_\nu}^{\delta^{18}O} &= \mu_{t_\nu}^{\delta^{18}O} + \eta_{j,t_\nu}^{\delta^{18}O} \\
\mu_{t_\nu + \Delta t_\nu}^{\delta^{13}C} &= \mu_{t_\nu}^{\delta^{13}C} + \eta_{j,t_\nu}^{\delta^{13}C}
\end{align*}
$$
The error terms of the unobserved components in $\delta^{18}$O and the one in $\delta^{13}$C correlate with coefficient $\rho_j$: 
$$
E[(\eta_{j,t_\nu}^{\delta^{18}O}\eta_{j,t_\nu}^{\delta^{13}C})'(\eta_{j,t_\nu}^{\delta^{18}O}\eta_{j,t_\nu}^{\delta^{13}C})] = Q_{t_\nu} = \left[\begin{array}{cc}
\sigma_{\eta,\delta^{18}O,j}^2 \Delta t_\nu^{\delta^{18}O} & \rho_j \sigma_{\eta,\delta^{18}O,j}\sigma_{\eta,\delta^{13}C,j} \min\{\Delta t_\nu^{\delta^{18}O}, \Delta t_\nu^{\delta^{13}C}\} \\
\rho_j\sigma_{\eta,\delta^{18}O,j}\sigma_{\eta,\delta^{13}C,j} \min\{\Delta t_\nu^{\delta^{18}O}, \Delta t_\nu^{\delta^{13}C}\} & \sigma_{\eta,\delta^{13}C,j}^2 \Delta t_\nu^{\delta^{13}C}
\end{array}\right],
$$
where the $j=1,\ldots,6$ indicate the climate state, and
$$
\Delta t_\nu^{(\delta^{18}O,\delta^{13}C)} = \left\{\begin{array}{l}
\Delta t_\nu,  \qquad\qquad\quad \text{if preceding obs at $t_{\nu-1}$ is not NA},\\
\Delta t_\nu + \Delta t_{\nu-1}^{(\delta^{18}O,\delta^{13}C)}, \text{if preceding obs at $t_{\nu-1}$ is NA}.
\end{array}\right.
$$


```{r, Model 4, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
if (Model==4){
  mObs<-mData[,6:13] # Delta 18O (6:9) Delta 13C (10:13)
  mZ<-matrix(0,8,2)
  mZ[1:4,1]<-1
  mZ[5:8,2]<-1
  
  mDeltaTau<-array(0,c(2,2,cT))
  vDeltaTau1<-matrix(0,cT,1)
  vDeltaTau2<-matrix(0,cT,1)
  DeltaTau1<-0
  DeltaTau2<-0
  for (i in 1:cT){
    if (is.na(mData[i,6])){ # d18O
      DeltaTau1<-DeltaTau1+mData[i,5]
    } else {
      DeltaTau1<-DeltaTau1+mData[i,5]
      vDeltaTau1[i]<-DeltaTau1
      DeltaTau1<-0
    }
    if (is.na(mData[i,10])){ # d13C
      DeltaTau2<-DeltaTau2+mData[i,5]
    } else {
      DeltaTau2<-DeltaTau2+mData[i,5]
      vDeltaTau2[i]<-DeltaTau2
      DeltaTau2<-0
    }
  }
  mDeltaTau[1,1,]<-vDeltaTau1
  mDeltaTau[2,2,]<-vDeltaTau2
  for (i in 1:cT){
    mDeltaTau[1,2,i]<-mDeltaTau[2,1,i]<-min(vDeltaTau1[i],vDeltaTau2[i])
  }
  mDeltaTau[, ,cT]<-matrix(0,2,2)
  
  mdl<-SSModel(as.matrix(mObs) ~ -1 + SSMcustom(Z=mZ,T=diag(2),Q=array(NA,c(2,2,cT)),a1=matrix(0,2,1),P1=matrix(0,2,2),P1inf=diag(2)),H=array(NA,c(8,8,cT)))
  
  # objective function of the state space model
  objf <- function(pars,model,estimate=TRUE){
    mQ<-array(0,c(2,2,cT))
    mQpars<-cbind(exp(pars[69:74]),exp(pars[75:80]),(1-exp(-pars[81:86]))/(1+exp(-pars[81:86])))
    mQ[1,1,1:2690]<-mQpars[1,1]*mDeltaTau[1,1,1:2690]
    mQ[2,2,1:2690]<-mQpars[1,2]*mDeltaTau[2,2,1:2690]
    mQ[1,2,1:2690]<-mQ[2,1,1:2690]<-mQpars[1,3]*sqrt(mQpars[1,1])*sqrt(mQpars[1,2])*mDeltaTau[1,2,1:2690]
    mQ[1,1,2691:5693]<-mQpars[2,1]*mDeltaTau[1,1,2691:5693]
    mQ[2,2,2691:5693]<-mQpars[2,2]*mDeltaTau[2,2,2691:5693]
    mQ[1,2,2691:5693]<-mQ[2,1,2691:5693]<-mQpars[2,3]*sqrt(mQpars[2,1])*sqrt(mQpars[2,2])*mDeltaTau[1,2,2691:5693]
    mQ[1,1,5694:7471]<-mQpars[3,1]*mDeltaTau[1,1,5694:7471]
    mQ[2,2,5694:7471]<-mQpars[3,2]*mDeltaTau[2,2,5694:7471]
    mQ[1,2,5694:7471]<-mQ[2,1,5694:7471]<-mQpars[3,3]*sqrt(mQpars[3,1])*sqrt(mQpars[3,2])*mDeltaTau[1,2,5694:7471]
    mQ[1,1,7472:13933]<-mQpars[4,1]*mDeltaTau[1,1,7472:13933]
    mQ[2,2,7472:13933]<-mQpars[4,2]*mDeltaTau[2,2,7472:13933]
    mQ[1,2,7472:13933]<-mQ[2,1,7472:13933]<-mQpars[4,3]*sqrt(mQpars[4,1])*sqrt(mQpars[4,2])*mDeltaTau[1,2,7472:13933]
    mQ[1,1,13934:20179]<-mQpars[5,1]*mDeltaTau[1,1,13934:20179]
    mQ[2,2,13934:20179]<-mQpars[5,2]*mDeltaTau[2,2,13934:20179]
    mQ[1,2,13934:20179]<-mQ[2,1,13934:20179]<-mQpars[5,3]*sqrt(mQpars[5,1])*sqrt(mQpars[5,2])*mDeltaTau[1,2,13934:20179]
    mQ[1,1,20180:23722]<-mQpars[6,1]*mDeltaTau[1,1,20180:23722]
    mQ[2,2,20180:23722]<-mQpars[6,2]*mDeltaTau[2,2,20180:23722]
    mQ[1,2,20180:23722]<-mQ[2,1,20180:23722]<-mQpars[6,3]*sqrt(mQpars[6,1])*sqrt(mQpars[6,2])*mDeltaTau[1,2,20180:23722]
    model$Q<-mQ
    mH=array(0,c(8,8,cT))
    mHpars<-cbind(exp(pars[1:34]),exp(pars[35:68]))
    for (i in 1:cT){
      idx1<-which(!is.na(mData[i,14:17]))
      mmH1<-matrix(0,4,4)
      vDiag1<-matrix(0,4,1)
      vDiag1[idx1]<-mHpars[mData[i,13+idx1],1]
      mmH1[row(mmH1)==col(mmH1)]<-vDiag1
      mH[1:4,1:4,i]<-mmH1
      idx2<-which(!is.na(mData[i,18:21]))
      mmH2<-matrix(0,4,4)
      vDiag2<-matrix(0,4,1)
      vDiag2[idx2]<-mHpars[mData[i,17+idx2],2]
      mmH2[row(mmH2)==col(mmH2)]<-vDiag2
      mH[5:8,5:8,i]<-mmH2
    }
    model$H<-mH
    if (estimate){
      -logLik(model)
    } else {
      model
    }
  }
  
  # estimation
  github_url<-"https://github.com/chisos9/data_public/raw/master/Bivariate_locallevel_studies_periods_par.rda"
  load(url(github_url)) # load parameter start vector
  mdl.opt<-optim(vPi,fn=objf,model=mdl,method="BFGS",control=list(trace=4,maxit=10000),hessian=FALSE) 
  vPi<-mdl.opt$par

  # evaluate Kalman filter and smoother in the optimum
  mdl.fit<-objf(vPi, mdl, estimate = FALSE)
  mdl.kfs<-KFS(mdl.fit,filtering=c('state','signal','disturbance'),smoothing=c('state','signal','disturbance'),simplify=FALSE)
  
  # Generate equidistantly time-stamped data
  timeax<-seq(from=-67,to=0,by=(Delta/1e6))
  cT_gen<-length(timeax)
  mData_equi<-matrix(NA,cT_gen,29)
  mData_equi[,4]<-timeax
  mData_merged<-rbind(mData,mData_equi)
  indices_in_timestamps_order<-order(mData_merged[,4])
  mData_gen<-mData_merged[indices_in_timestamps_order,]
  indices_equi<-matrix(NA,cT_gen,1)
  for (i in 1:cT_gen){
    indices_equi[i]<-which(indices_in_timestamps_order==(cT+i))
  }
  Dtau_SSM_gen<-c(diff(mData_gen[,4]),NA)
  mDeltaTau_gen<-array(0,c(2,2,cT+cT_gen))
  vDeltaTau1_gen<-matrix(0,cT+cT_gen,1)
  vDeltaTau2_gen<-matrix(0,cT+cT_gen,1)
  DeltaTau1<-0
  DeltaTau2<-0
  for (i in 1:(cT+cT_gen)){
    if (is.na(mData_gen[i,6])){ # d18O
      DeltaTau1<-DeltaTau1+Dtau_SSM_gen[i]
    } else {
      DeltaTau1<-DeltaTau1+Dtau_SSM_gen[i]
      vDeltaTau1_gen[i]<-DeltaTau1
      DeltaTau1<-0
    }
    if (is.na(mData_gen[i,10])){ # d13C
      DeltaTau2<-DeltaTau2+Dtau_SSM_gen[i]
    } else {
      DeltaTau2<-DeltaTau2+Dtau_SSM_gen[i]
      vDeltaTau2_gen[i]<-DeltaTau2
      DeltaTau2<-0
    }
  }
  mDeltaTau_gen[1,1,]<-vDeltaTau1_gen
  mDeltaTau_gen[2,2,]<-vDeltaTau2_gen
  for (i in 1:(cT+cT_gen)){
    mDeltaTau_gen[1,2,i]<-mDeltaTau_gen[2,1,i]<-min(vDeltaTau1_gen[i],vDeltaTau2_gen[i])
  }
  mDeltaTau_gen[, ,cT+cT_gen]<-matrix(0,2,2)
  mQ<-array(0,c(2,2,cT+cT_gen))
  mH=array(0,c(8,8,cT+cT_gen))
  # objective function of the state space model
  objf_gen <- function(pars,model,estimate=TRUE){
    mQpars<-cbind(exp(pars[69:74]),exp(pars[75:80]),(1-exp(-pars[81:86]))/(1+exp(-pars[81:86])))
    for (i in 1:(cT+cT_gen)){
      if (mData_gen[i,4]< (-56)){                                             
        mQ[1,1,i]<-mQpars[1,1]*mDeltaTau_gen[1,1,i]
        mQ[2,2,i]<-mQpars[1,2]*mDeltaTau_gen[2,2,i]
        mQ[1,2,i]<-mQ[2,1,i]<-mQpars[1,3]*sqrt(mQpars[1,1])*sqrt(mQpars[1,2])*mDeltaTau_gen[1,2,i]
      } else if ((mData_gen[i,4] >= (-56)) && (mData_gen[i,4] < (-47))) {     
        mQ[1,1,i]<-mQpars[2,1]*mDeltaTau_gen[1,1,i]
        mQ[2,2,i]<-mQpars[2,2]*mDeltaTau_gen[2,2,i]
        mQ[1,2,i]<-mQ[2,1,i]<-mQpars[2,3]*sqrt(mQpars[2,1])*sqrt(mQpars[2,2])*mDeltaTau_gen[1,2,i]
      } else if ((mData_gen[i,4] >= (-47)) && (mData_gen[i,4] < (-34))) {     
        mQ[1,1,i]<-mQpars[3,1]*mDeltaTau_gen[1,1,i]
        mQ[2,2,i]<-mQpars[3,2]*mDeltaTau_gen[2,2,i]
        mQ[1,2,i]<-mQ[2,1,i]<-mQpars[3,3]*sqrt(mQpars[3,1])*sqrt(mQpars[3,2])*mDeltaTau_gen[1,2,i]
      } else if ((mData_gen[i,4] >= (-34)) && (mData_gen[i,4] < (-13.9))) {   
        mQ[1,1,i]<-mQpars[4,1]*mDeltaTau_gen[1,1,i]
        mQ[2,2,i]<-mQpars[4,2]*mDeltaTau_gen[2,2,i]
        mQ[1,2,i]<-mQ[2,1,i]<-mQpars[4,3]*sqrt(mQpars[4,1])*sqrt(mQpars[4,2])*mDeltaTau_gen[1,2,i]
      } else if ((mData_gen[i,4] >= (-13.9)) && (mData_gen[i,4] < (-3.3))) {
        mQ[1,1,i]<-mQpars[5,1]*mDeltaTau_gen[1,1,i]
        mQ[2,2,i]<-mQpars[5,2]*mDeltaTau_gen[2,2,i]
        mQ[1,2,i]<-mQ[2,1,i]<-mQpars[5,3]*sqrt(mQpars[5,1])*sqrt(mQpars[5,2])*mDeltaTau_gen[1,2,i]
      } else if (mData_gen[i,4] >= (-3.3)) {
        mQ[1,1,i]<-mQpars[6,1]*mDeltaTau_gen[1,1,i]
        mQ[2,2,i]<-mQpars[6,2]*mDeltaTau_gen[2,2,i]
        mQ[1,2,i]<-mQ[2,1,i]<-mQpars[6,3]*sqrt(mQpars[6,1])*sqrt(mQpars[6,2])*mDeltaTau_gen[1,2,i]
      }
    }
    model$Q<-mQ
    mHpars<-cbind(exp(pars[1:34]),exp(pars[35:68]))
    for (i in 1:(cT+cT_gen)){
      idx1<-which(!is.na(mData_gen[i,14:17]))
      mmH1<-matrix(0,4,4)
      vDiag1<-matrix(0,4,1)
      vDiag1[idx1]<-mHpars[mData_gen[i,13+idx1],1]
      mmH1[row(mmH1)==col(mmH1)]<-vDiag1
      mH[1:4,1:4,i]<-mmH1
      idx2<-which(!is.na(mData_gen[i,18:21]))
      mmH2<-matrix(0,4,4)
      vDiag2<-matrix(0,4,1)
      vDiag2[idx2]<-mHpars[mData_gen[i,17+idx2],2]
      mmH2[row(mmH2)==col(mmH2)]<-vDiag2
      mH[5:8,5:8,i]<-mmH2
    }
    model$H<-mH
    if (estimate){
      -logLik(model)
    } else {
      model
    }
  }
  mdl_gen<-SSModel(mData_gen[,6:13] ~ -1 + SSMcustom(Z=mZ,T=diag(2),Q=array(NA,c(2,2,cT+cT_gen)),a1=matrix(0,2,1),P1=matrix(0,2,2),P1inf=diag(2)),H=array(NA,c(8,8,cT+cT_gen)))
  mdl.fit_gen<-objf_gen(vPi, mdl_gen, estimate = FALSE)
  mdl.kfs_gen<-KFS(mdl.fit_gen,filtering=c('state','signal','disturbance'),smoothing=c('state','signal','disturbance'),simplify=FALSE)
  output_series<-mdl.kfs_gen$alphahat[indices_equi,]
  output_series_d18O<-round(output_series[,1],digits=10)
  output_series_d13C<-round(output_series[,2],digits=10)
  output_matrix<-cbind(round(timeax,digits=6),output_series_d18O,output_series_d13C)
  write.table(output_matrix,file="output.txt",sep="\t",row.names=FALSE)
}
```


## Model 5: Integrated random walk plus noise for integration orders 2 to 6.

In this model, the unobserved component is modeled as an integrated random walk of order $m$. The transition equation of this class of models is specified as follows.

$$
\left[\begin{array}{c}
\mu_{t_\nu+\Delta t_\nu} \\ \mu_{t_\nu+\Delta t_\nu}^{(m-1)} \\ \vdots \\ \mu_{t_\nu+\Delta t_\nu}^{(2)} \\ \mu_{t_\nu+\Delta t_\nu}^{(1)}
\end{array}\right]
=
\left[\begin{array}{ccccccc}
1 & 1 & 0 & 0 & \cdots & 0 & 0 \\ 
0 & 1 & 1 & 0 & \cdots & 0 & 0 \\ 
\vdots & \vdots & \vdots & \vdots & \ddots & \vdots & \vdots \\
0 & 0 & 0 & 0 & \cdots & 1 & 1 \\
0 & 0 & 0 & 0 & \cdots & 0 & 1
\end{array}\right]
\left[\begin{array}{c}
\mu_{t_\nu} \\ \mu_{t_\nu}^{(m-1)} \\ \vdots \\ \mu_{t_\nu}^{(2)} \\ \mu_{t_\nu}^{(1)}
\end{array}\right]
+
\left[\begin{array}{c}
0 \\ 0 \\ \vdots \\ 0 \\ \eta_{j,t_\nu}
\end{array}\right],\; \eta_{j,t_\nu}\stackrel{i.i.d.}{\sim} \mathsf{N}(0, \sigma_{\eta,j,m}^2 \Delta t_\nu).
$$
where $j=1,\ldots,6$ denotes climate states. Here, $\mu^{(1)}$ is a random walk with white noise driver $\eta$ that gets cumulated $m$ times without adding additional sources of noise in the $m-1$ cumulations. As a result, $\mu$ is an $m$-fold integrated random walk. This is an $I(m)$ process, meaning that $m$-fold differencing yields a stationary process. In the paper, we discuss how the order of integration relates to the order of a Butterworth filter. The parameter $m$ set above is only used in this model, and it can currently be set to a natural number between 2 and 6. Note that $m=1$ is Model 3. The measurement equation is defined as in Model 2 and distinguishes study variances.


```{r, Model 5, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
if (Model==5){
  mObs_d18O<-mData[,6:9] # Delta 18O
  mObs_d13C<-mData[,10:13] # Delta 13C
  mZ<-matrix(1,4,m)
  mT<-diag(m)
  mT[(row(mT)==col(mT)) | (col(mT)==row(mT)+1)]<-1
  vDeltaTau_d18O<-matrix(0,cT,1)
  vDeltaTau_d13C<-matrix(0,cT,1)
  DeltaTau<-0
  for (i in 1:cT){
    if (is.na(mData[i,6])){
      DeltaTau<-DeltaTau+mData[i,5]
    } else {
      DeltaTau<-DeltaTau+mData[i,5]
      vDeltaTau_d18O[i]<-DeltaTau
      DeltaTau<-0
    }
  }
  DeltaTau<-0
  for (i in 1:cT){
    if (is.na(mData[i,10])){
      DeltaTau<-DeltaTau+mData[i,5]
    } else {
      DeltaTau<-DeltaTau+mData[i,5]
      vDeltaTau_d13C[i]<-DeltaTau
      DeltaTau<-0
    }
  }
  mR_d18O<-array(0,c(m,m,cT))
  mR_d13C<-array(0,c(m,m,cT))
  mR_d18O[m,m,]<-sqrt(vDeltaTau_d18O)
  mR_d13C[m,m,]<-sqrt(vDeltaTau_d13C)
  mR_d18O[, ,cT]<-0
  mR_d13C[, ,cT]<-0
  mdl_d18O<-SSModel(as.matrix(mObs_d18O) ~ -1 + SSMcustom(Z=mZ,T=mT,R=mR_d18O,Q=array(NA,c(m,m,cT)),a1=matrix(0,m,1),P1=diag(0,m),P1inf=diag(m)),H=array(NA,c(4,4,cT)))
  mdl_d13C<-SSModel(as.matrix(mObs_d13C) ~ -1 + SSMcustom(Z=mZ,T=mT,R=mR_d13C,Q=array(NA,c(m,m,cT)),a1=matrix(0,m,1),P1=diag(0,m),P1inf=diag(m)),H=array(NA,c(4,4,cT)))
  
  # objective function of the state space model
  objf_d18O <- function(pars,model,estimate=TRUE){
    mQ<-array(0,c(m,m,cT))
    mQ[m,m,1:2690]<-exp(pars[35])
    mQ[m,m,2691:5693]<-exp(pars[36])
    mQ[m,m,5694:7471]<-exp(pars[37])
    mQ[m,m,7472:13933]<-exp(pars[38])
    mQ[m,m,13934:20179]<-exp(pars[39])
    mQ[m,m,20180:23722]<-exp(pars[40])
    model$Q<-mQ
    mH=array(0,c(4,4,cT))
    for (i in 1:cT){
      idx<-which(!is.na(mData[i,14:17]))
      mmH<-matrix(0,4,4)
      vDiag<-matrix(0,4,1)
      vDiag[idx]<-exp(pars[mData[i,13+idx]])
      mmH[row(mmH)==col(mmH)]<-vDiag
      mH[, ,i]<-mmH
    }
    model$H<-mH
    if (estimate){
      -logLik(model)
    } else {
      model
    }
  }
  objf_d13C <- function(pars,model,estimate=TRUE){
    mQ<-array(0,c(m,m,cT))
    mQ[m,m,1:2690]<-exp(pars[35])
    mQ[m,m,2691:5693]<-exp(pars[36])
    mQ[m,m,5694:7471]<-exp(pars[37])
    mQ[m,m,7472:13933]<-exp(pars[38])
    mQ[m,m,13934:20179]<-exp(pars[39])
    mQ[m,m,20180:23722]<-exp(pars[40])
    model$Q<-mQ
    mH=array(0,c(4,4,cT))
    for (i in 1:cT){
      idx<-which(!is.na(mData[i,18:21]))
      mmH<-matrix(0,4,4)
      vDiag<-matrix(0,4,1)
      vDiag[idx]<-exp(pars[mData[i,17+idx]])
      mmH[row(mmH)==col(mmH)]<-vDiag
      mH[, ,i]<-mmH
    }
    model$H<-mH
    if (estimate){
      -logLik(model)
    } else {
      model
    }
  }
  # estimation
  if (m==2){
    github_url<-"https://github.com/chisos9/data_public/raw/master/Delta18O_localtrend_studies_periods_par.rda"
    load(url(github_url)) # load parameter start vector
  } else if (m==3){
    github_url<-"https://github.com/chisos9/data_public/raw/master/Delta18O_localtrend_m3_studies_periods_par.rda"
    load(url(github_url)) # load parameter start vector
  } else if (m==4){
    github_url<-"https://github.com/chisos9/data_public/raw/master/Delta18O_localtrend_m4_studies_periods_par.rda"
    load(url(github_url)) # load parameter start vector
  } else if (m==5){
    github_url<-"https://github.com/chisos9/data_public/raw/master/Delta18O_localtrend_m5_studies_periods_par.rda"
    load(url(github_url)) # load parameter start vector
  } else if (m==6){
    github_url<-"https://github.com/chisos9/data_public/raw/master/Delta18O_localtrend_m6_studies_periods_par.rda"
    load(url(github_url)) # load parameter start vector
  }
  mdl.opt_d18O<-optim(vPi,fn=objf_d18O,model=mdl_d18O,method="BFGS",control=list(trace=4,maxit=10000),hessian=FALSE) # use your favorite numerical optimizer here
  vPi_d18O<-mdl.opt_d18O$par
  
  # evaluate Kalman filter and smoother in the optimum
  mdl.fit_d18O<-objf_d18O(vPi_d18O, mdl_d18O, estimate = FALSE)
  mdl.kfs_d18O<-KFS(mdl.fit_d18O,filtering=c('state','signal','disturbance'),smoothing=c('state','signal','disturbance'),simplify=FALSE)
  
  if (m==2){
    github_url<-"https://github.com/chisos9/data_public/raw/master/Delta13C_localtrend_studies_periods_par.rda"
    load(url(github_url)) # load parameter start vector
  } else if (m==3){
    github_url<-"https://github.com/chisos9/data_public/raw/master/Delta13C_localtrend_m3_studies_periods_par.rda"
    load(url(github_url)) # load parameter start vector
  } else if (m==4){
    github_url<-"https://github.com/chisos9/data_public/raw/master/Delta13C_localtrend_m4_studies_periods_par.rda"
    load(url(github_url)) # load parameter start vector
  } else if (m==5){
    github_url<-"https://github.com/chisos9/data_public/raw/master/Delta13C_localtrend_m5_studies_periods_par.rda"
    load(url(github_url)) # load parameter start vector
  } else if (m==6){
    github_url<-"https://github.com/chisos9/data_public/raw/master/Delta13C_localtrend_m6_studies_periods_par.rda"
    load(url(github_url)) # load parameter start vector
  }
  mdl.opt_d13C<-optim(vPi,fn=objf_d13C,model=mdl_d13C,method="BFGS",control=list(trace=4,maxit=10000),hessian=FALSE) # use your favorite numerical optimizer here
  vPi_d13C<-mdl.opt_d13C$par
  
  mdl.fit_d13C<-objf_d13C(vPi_d13C, mdl_d13C, estimate = FALSE)
  mdl.kfs_d13C<-KFS(mdl.fit_d13C,filtering=c('state','signal','disturbance'),smoothing=c('state','signal','disturbance'),simplify=FALSE)
  
  timeax<-seq(from=-67,to=0,by=(Delta/1e6))
  cT_gen<-length(timeax)
  mData_equi<-matrix(NA,cT_gen,29)
  mData_equi[,4]<-timeax
  mData_merged<-rbind(mData,mData_equi)
  indices_in_timestamps_order<-order(mData_merged[,4])
  mData_gen<-mData_merged[indices_in_timestamps_order,]
  indices_equi<-matrix(NA,cT_gen,1)
  for (i in 1:cT_gen){
    indices_equi[i]<-which(indices_in_timestamps_order==(cT+i))
  }
  Dtau_SSM_gen<-c(diff(mData_gen[,4]),NA)
  vDeltaTau_gen_d18O<-matrix(0,cT+cT_gen,1)
  DeltaTau<-0
  for (i in 1:(cT+cT_gen)){
    if (is.na(mData_gen[i,6])){
      DeltaTau<-DeltaTau+Dtau_SSM_gen[i]
    } else {
      DeltaTau<-DeltaTau+Dtau_SSM_gen[i]
      vDeltaTau_gen_d18O[i]<-DeltaTau
      DeltaTau<-0
    }
  }
  vDeltaTau_gen_d13C<-matrix(0,cT+cT_gen,1)
  DeltaTau<-0
  for (i in 1:(cT+cT_gen)){
    if (is.na(mData_gen[i,10])){
      DeltaTau<-DeltaTau+Dtau_SSM_gen[i]
    } else {
      DeltaTau<-DeltaTau+Dtau_SSM_gen[i]
      vDeltaTau_gen_d13C[i]<-DeltaTau
      DeltaTau<-0
    }
  }
  mR_gen_d18O<-array(0,c(m,m,cT+cT_gen))
  mR_gen_d13C<-array(0,c(m,m,cT+cT_gen))
  mR_gen_d18O[m,m,]<-sqrt(vDeltaTau_gen_d18O)
  mR_gen_d13C[m,m,]<-sqrt(vDeltaTau_gen_d13C)
  mR_gen_d18O[ , ,cT+cT_gen]<-0
  mR_gen_d13C[ , ,cT+cT_gen]<-0
  # objective function of the state space model
  objf_gen_d18O <- function(pars,model,estimate=TRUE){
    mQ<-array(0,c(m,m,cT+cT_gen))
    for (i in 1:(cT+cT_gen)){
      if (mData_gen[i,4]< (-56)){                                             
        mQ[m,m,i]<-exp(pars[35])
      } else if ((mData_gen[i,4] >= (-56)) && (mData_gen[i,4] < (-47))) {     
        mQ[m,m,i]<-exp(pars[36])
      } else if ((mData_gen[i,4] >= (-47)) && (mData_gen[i,4] < (-34))) {     
        mQ[m,m,i]<-exp(pars[37])
      } else if ((mData_gen[i,4] >= (-34)) && (mData_gen[i,4] < (-13.9))) {   
        mQ[m,m,i]<-exp(pars[38])
      } else if ((mData_gen[i,4] >= (-13.9)) && (mData_gen[i,4] < (-3.3))) {
        mQ[m,m,i]<-exp(pars[39])
      } else if (mData_gen[i,4] >= (-3.3)) {
        mQ[m,m,i]<-exp(pars[40])
      }
    }
    model$Q<-mQ
    mH<-array(0,c(4,4,cT+cT_gen))
    for (i in 1:(cT+cT_gen)){
      idx<-which(!is.na(mData_gen[i,14:17]))
      mmH<-matrix(0,4,4)
      vDiag<-matrix(0,4,1)
      vDiag[idx]<-exp(pars[mData_gen[i,13+idx]])
      mmH[row(mmH)==col(mmH)]<-vDiag
      mH[, ,i]<-mmH
    }
    model$H<-mH
    if (estimate){
      -logLik(model)
    } else {
      model
    }
  }
  objf_gen_d13C <- function(pars,model,estimate=TRUE){
    mQ<-array(0,c(m,m,cT+cT_gen))
    for (i in 1:(cT+cT_gen)){
      if (mData_gen[i,4]< (-56)){                                             
        mQ[m,m,i]<-exp(pars[35])
      } else if ((mData_gen[i,4] >= (-56)) && (mData_gen[i,4] < (-47))) {     
        mQ[m,m,i]<-exp(pars[36])
      } else if ((mData_gen[i,4] >= (-47)) && (mData_gen[i,4] < (-34))) {     
        mQ[m,m,i]<-exp(pars[37])
      } else if ((mData_gen[i,4] >= (-34)) && (mData_gen[i,4] < (-13.9))) {   
        mQ[m,m,i]<-exp(pars[38])
      } else if ((mData_gen[i,4] >= (-13.9)) && (mData_gen[i,4] < (-3.3))) {
        mQ[m,m,i]<-exp(pars[39])
      } else if (mData_gen[i,4] >= (-3.3)) {
        mQ[m,m,i]<-exp(pars[40])
      }
    }
    model$Q<-mQ
    mH<-array(0,c(4,4,cT+cT_gen))
    for (i in 1:(cT+cT_gen)){
      idx<-which(!is.na(mData_gen[i,18:21]))
      mmH<-matrix(0,4,4)
      vDiag<-matrix(0,4,1)
      vDiag[idx]<-exp(pars[mData_gen[i,17+idx]])
      mmH[row(mmH)==col(mmH)]<-vDiag
      mH[, ,i]<-mmH
    }
    model$H<-mH
    if (estimate){
      -logLik(model)
    } else {
      model
    }
  }
  mdl_gen_d18O<-SSModel(mData_gen[,6:9] ~ -1 + SSMcustom(Z=mZ,T=mT,R=mR_gen_d18O,Q=array(NA,c(m,m,cT+cT_gen)),a1=matrix(0,m,1),P1=diag(0,m),P1inf=diag(m)),H=array(NA,c(4,4,cT+cT_gen)))
  mdl_gen_d13C<-SSModel(mData_gen[,10:13] ~ -1 + SSMcustom(Z=mZ,T=mT,R=mR_gen_d13C,Q=array(NA,c(m,m,cT+cT_gen)),a1=matrix(0,m,1),P1=diag(0,m),P1inf=diag(m)),H=array(NA,c(4,4,cT+cT_gen)))
  mdl.fit_gen_d18O<-objf_gen_d18O(vPi_d18O, mdl_gen_d18O, estimate = FALSE)
  mdl.fit_gen_d13C<-objf_gen_d13C(vPi_d13C, mdl_gen_d13C, estimate = FALSE)
  mdl.kfs_gen_d18O<-KFS(mdl.fit_gen_d18O,filtering=c('state','signal','disturbance'),smoothing=c('state','signal','disturbance'),simplify=FALSE)
  mdl.kfs_gen_d13C<-KFS(mdl.fit_gen_d13C,filtering=c('state','signal','disturbance'),smoothing=c('state','signal','disturbance'),simplify=FALSE)
  output_series_d18O<-round(mdl.kfs_gen_d18O$alphahat[indices_equi],digits=10)
  output_series_d13C<-round(mdl.kfs_gen_d13C$alphahat[indices_equi],digits=10)
  output_matrix<-cbind(round(timeax,digits=6),output_series_d18O,output_series_d13C)
  write.table(output_matrix,file="output.txt",sep="\t",row.names=FALSE)
}
```

## Regularly time-stamped smoothed values

The plot shows equidistantly time-stamped smoothed values from the Kalman smoothing recursions for the chosen model and the chosen time increment $\Delta$ for $\delta^{18}$O (left panel) and $\delta^{13}$C (right panel).

```{r, Final plots, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10}
# time series plot
par(mfrow = c(1, 2))

plot(mData[,4],mData[,6],type='l',lwd=.2,col='gray',ylim=rev(range(mData[,6],na.rm=TRUE)),ylab='Delta-O-18',xlab='Million years before present')
lines(timeax,output_series_d18O,ylim=rev(range(output_series_d18O,na.rm=TRUE)),lwd=1,col='black')
legend("topright",c("data","regular smoothed values"),col=c("gray","black"),lty=c(1,1),lwd=c(2,2),bty='n')
  
plot(mData[,4],mData[,10],type='l',lwd=.2,col='gray',ylim=range(mData[,10],na.rm=TRUE),ylab='Delta-C-13',xlab='Million years before present')
lines(timeax,output_series_d13C,ylim=range(output_series_d13C,na.rm=TRUE),lwd=1,col='black')
legend("topright",c("data","regular smoothed values"),col=c("gray","black"),lty=c(1,1),lwd=c(2,2),bty='n')


```

Regularly time-stamped imputed smoothed values from Model 4 for different $\Delta$'s can be downloaded at the links below.

* [$\Delta=1000$ Years](https://raw.githubusercontent.com/chisos9/data_public/master/output_model4_delta1000.txt) 
* [$\Delta=5000$ Years](https://raw.githubusercontent.com/chisos9/data_public/master/output_model4_delta5000.txt) 
* [$\Delta=10000$ Years](https://raw.githubusercontent.com/chisos9/data_public/master/output_model4_delta10000.txt) 
* [$\Delta=50000$ Years](https://raw.githubusercontent.com/chisos9/data_public/master/output_model4_delta50000.txt) 
* [$\Delta=100000$ Years](https://raw.githubusercontent.com/chisos9/data_public/master/output_model4_delta100000.txt) 
* Download this R-Markdown script (Available after publication of paper)



